---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Elegant Data Manipulation with Lenses

## Installation

```{r, eval = FALSE}
devtools::install_github("cfhammill/lenses")
```

## Intro

In typical R use we frequently perform two common operations on our data.
We `view` some piece of the data, or we `set` some piece of the data. Base
R comes with many pairs of `view` and `set` functions. A classic example
would be `names`. Names can be viewed `names(x)` and set `names(x) <- new_names`.
Lenses take this idea and offer some important advantages:

- bidirectionality (you can `set` anything you can `view`)
- composability (lenses can be put together to `set`/`view` nested data)
- reduces code duplication
- non-destructive (lenses won't update your data without your permission)

In this document, we'll see a few common data manipulation operations and
how they can be improved with lenses.

## Simple manipulations

Let's take the iris data set for example, we want to perform
some manipulations on it.

```{r, message = FALSE}
library(lenses)
library(dplyr)

head(iris)
```

We're curious about the value of the 3rd element of the `Sepal.Length` column.
Using base R we can `view` it with: 

```{r}
iris$Sepal.Length[3]
```

we can update (set) the value by assigning into it:

```{r}
iris$Sepal.Length[3] <- 100
head(iris, 3)
```

and we can perform some operation to update it:

```{r}
iris$Sepal.Length[3] <- log(iris$Sepal.Length[3])
```

This works well, however, there are some problems. 

The first problem is comes with having our  `view` and `set` functions separate. 
Composing our operations isn't easy, particularly when using pipes:

```{r}
iris %>%
  .$Sepal.Length %>%
  `[<-`(20)
```

Whoops, that's not what we wanted.

R's `set`/`view` pairs also can't be composed with function 
compostion e.g. `purrr::compose`. 

This is a failure of "bidirectionality", once you've chosen to use a 
`view` function, or a `set` function, you are locked into that direction.

Lack of composability and bidirectionality means that you frequently have
to duplicate your code. For example, if you want to apply an operation to the
third element of "Sepal.Length", you need to specify the chain of accessors
twice, once in `view` mode, and once in `set` mode.

```{r}
iris$Sepal.Length[3] <- iris$Sepal.Length[3] * 2
head(iris, 3)
```

We can fix both of these problems by using lenses.

## Using lenses

Lenses give you all the power of R's `view` and `set` functions plus
the advantages noted above. Especially important are the composition
and bidirectionality features. Each lens can be used with our
[`view`](https://cfhammill.github.io/lenses/reference/view.html),
and [`set`](https://cfhammill.github.io/lenses/reference/set.html).

Let's revisit the operations we performed above using lenses. 

The first thing we will do is construct a lens into the 
third element of `Sepal.Length`:

```{r}
sepal_length3 <- index("Sepal.Length") %.% index(3)
```

In the code above we're first creating a two lenses, one
into `Sepal.Length` and another into element 3 using the 
[`index`](https://cfhammill.github.io/lenses/reference/set.html) 
function. Then we're composing these two lenses with
[`%.%`](https://cfhammill.github.io/lenses/reference/lens-compose.html).

Note that this lens has no idea we're going to apply it
to `iris`. Lenses are constructed without view of the 
data.

Now that we have a lens into an the third element of `Sepal.Length`, 
we can examine the element with the 
[`view`](https://cfhammill.github.io/lenses/reference/view.html) function:

```{r}
iris %>% view(sepal_length3)
```

We can update this element with the 
[`set`](https://cfhammill.github.io/lenses/reference/set.html) function:

```{r}
iris %>% set(sepal_length3, 50) %>% head(3)
```

And we can apply a function to change the data. To
do this we can apply a function 
[`over`](https://cfhammill.github.io/lenses/reference/over.html) 
the lens:

```{r}
iris %>% over(sepal_length3, log) %>% head(3)
```

Note that we never had to respecify what subpart
we wanted, the lens kept track for us. We saw that
the same lens can be used to both 
[`view`](https://cfhammill.github.io/lenses/reference/view.html) and 
[`set`](https://cfhammill.github.io/lenses/reference/set.html),
and that they can be composed easily with 
[`%.%`](https://cfhammill.github.io/lenses/reference/lens-compose.html).

## More interesting lenses

Now you have seen the main lens verbs and operations

1. [`view`](https://cfhammill.github.io/lenses/reference/view.html): 
  see the subpart of an object a lens is focussed on.
1. [`set`](https://cfhammill.github.io/lenses/reference/set.html): 
  set the subpart to a particular value, then return the
  whole object with the subpart updated.
1. [`over`](https://cfhammill.github.io/lenses/reference/over.html): 
  apply a function to the subpart, then return the 
  whole object with the subpart updated.
1. [`%.%`](https://cfhammill.github.io/lenses/reference/lens-compose.html): 
  compose two lenses to focus on a subpart of a subpart.

Now if all lenses had to offer was more composable indexing of vectors, 
you  might not be interested in integrating them into your workflows. 
But lenses can do a lot more than just pick and set elements in vectors. 

For example, this package provides lens-ified version of `dplyr::select`.
Unlike `select`, 
[`select_l`](https://cfhammill.github.io/lenses/reference/select_l.html) 
is bidirectional. This means you can 
[`set`](https://cfhammill.github.io/lenses/reference/set.html) the results 
of your selection.

let's select columns between `Sepal.Width` and `Petal.Width` and increment them by 10:

```{r}
iris %>%
  over(select_l(Sepal.Width:Petal.Width)
     , ~ . + 10
       ) %>%
  head(3)
```

Not only does [`select_l`](https://cfhammill.github.io/lenses/reference/select_l.html) 
create the appropriate lens for you with `dplyr::select` style column references, 
but [`over`](https://cfhammill.github.io/lenses/reference/over.html) allows us 
to declare anonymous functions like in `purrr`.

At this point I can imagine you saying, all this is very clear, but what good is it, 
I have `mutate`, well that is certainly true. But have you ever wanted to `set` 
or modify the results of a `filter`?  This is not super easy to do in the `dplyr` 
universe. But our lens-ified `filter`, 
[`filter_l`](https://cfhammill.github.io/lenses/reference/filter_l.html) does this 
with ease.

Let's set all "sepal" columns where the row number is less than three to 
zero. And for fun let's also change the column names to all upper case:

```{r}
iris %>%
  mutate(row_num = seq_len(n())) %>%
  set(filter_l(row_num < 3) %.%
      select_l(matches("Sepal"))
    , 0) %>%
  over(names_l, toupper) %>%
  head(3)
```

As you can see, lenses can be smoothly integrated into your `tidyverse`
workflows, as well as your base R workflows. Giving you the powers
of compositionality and bidirectionality to improve your code.

## Polishing your own

You can make a lens from scratch (!) by passing `view` and `set` functions to 
the [`lens`](https://cfhammill.github.io/lenses/reference/lens.html) constructor:

```{r}
first_l <- lens(view = function(d) d[[1]],
                set  = function(d, x) { d[[1]] <- x; d })
```

As you see, the `view` function must accept an element of data, while the `set`
function must accepts such an element as well as the new value of the subpart,
and return the new data in its entirety - thus achieving composability -
 without modifying the original.

In order to avoid unpleasant surprises or inconsistencies for users,
an author of a `lens` (via [`lens`](https://cfhammill.github.io/lenses/reference/lens.html)) 
should ensure it obeys the following rules (the "Lenz laws", here paraphrased 
from [a Haskell lens tutorial](www.schoolofhaskell.com/school/to-infinity-and-beyond/pick-of-the-week/a-little-lens-starter-tutorial)):

1. View-Set: If you `view` some data with a lens, and then
`set` the data with that value, you get the input data back.
2. Set-View: If you `set` a value with a lens,
then `view` that value with the same lens, you get back what you put in.
3. Set-Set: If you `set` a value into some data with a lens, and
then `set` another value with the same lens, it's the same as only
doing the second `set`.

"Lenses" which do not satisfy these properties should be documented accordingly.
By convention, the few such specimens in this library are suffixed by "_il" ("illegal lens").

## How do they work?

As you can see from the `lens` constructor, knowing how to implement `view` and `set`
for a lens turns out to be sufficient to implement the other verbs such as `over` and
- most importantly - lens composition (`%.%`). 

In our implementation, `lens`es are trivial. They simply store the provided
functions. A `lens` under the hood is a two element list with an 
element `set` and an element `view`.

## History of lens making

There is nothing particularly new about the lenses appearing here.
For a fairly comprehensive (and highly technical) history of lenses, see
[links here](https://github.com/ekmett/lens/wiki/History-of-Lenses) and
[this blog post](https://julesh.com/2018/08/16/lenses-for-philosophers/) .
